Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-06-21 paper.tex
Small fixes

\begin{abstract}
  We present \emph{Decoupled Fallback}, a work-stealing method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum in WebGPU native and approaches the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsection{Earlier Attempts at Portability}
Our first attempt at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}. Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of starvation because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.

@@

We confirmed this queue-like progression experimentally by using atomics to record and print the order in which workgroups exited their lookback phase. This 'ordered exit' behavior is notably different from that of a \emph{Memcpy} kernel, where all in-flight workgroups typically begin writing immediately after completing reads, often leading to high-concurrency, uncoordinated global writes. In contrast, \emph{DF}'s staggered exit pattern implicitly throttles the number of concurrent writers, reducing memory pressure and improving locality. Workgroups that exit in near-order are more likely to access adjacent memory regions, reducing page thrashing and avoiding premature eviction of active pages. On devices whose memory systems may be insufficient to support highly concurrent writes---such as the 7900XT and Mali-G78---these effects plausibly contribute to scan throughput that exceeds \emph{Memcpy}, despite the added arithmetic. To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a $\sim$10\% throughput increase on the 7900XT. These findings suggest that \emph{Chained Scan}'s computational structure not only enables progress without explicit synchronization, but also improves global memory efficiency by shaping the temporal pattern of writes, effectively translating computational work into emergent memory coordination.

@@

  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups. Ideally, this number would be zero on devices with FPG\@; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched. In practice, this count is highly dependent on the fairness of the hardware scheduler. On most devices tested, initiated fallbacks remain modest---about 10\% or fewer of the total workgroups launched.

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near-100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test is limited in that it only measures the cost of increased fallback operations rather than truly simulating blocking behavior. In true starvation, a blocking workgroup stalls before posting its reduction and cannot continue until it receives execution from the scheduler. By contrast, our simulation allows workgroups to proceed and eventually post their inclusive reductions to the spine. Thus, we would expect true starvation to result in increased lookback lengths---unlike the slight downward trend observed in our simulation. Lastly, starvation leads to effective \emph{occupancy reduction}, something we cannot simulate dynamically: execution resources remain available, but fewer subgroups are scheduled, reducing opportunities for context switching to mitigate latency.

>>>>

\begin{abstract}
  We present \emph{Decoupled Fallback}, a work-stealing method that enables single-pass \emph{Chained Scans} to run on hardware without \emph{forward-progress guarantees} (FPG) while avoiding \emph{starvation}. Additionally, we introduce a tile state representation for \emph{Chained Scans} that does not rely on 64-bit atomics or memory barriers for correctness, along with a subgroup-size-agnostic intra-workgroup implementation. On FPG-lacking devices---Apple M1 Max and M3---\emph{Decoupled Fallback} achieves near-\emph{Memcpy} speeds for inclusive prefix sum using the native \emph{Dawn} implementation of the WebGPU standard, and approaches the full theoretically expected 50\% speedup over the slower \emph{Reduce-then-Scan} approach. We further demonstrate the resilience of \emph{Decoupled Fallback} against \emph{unfair} schedulers by simulating blocking at rates of up to 50\%, showing that it maintains superior performance over \emph{Reduce-then-Scan} even under extreme contention.
\end{abstract}

@@

\subsection{Earlier Attempts at Portability}
Our first attempt at implementing a \emph{Decoupled Lookback} scan without relying on FPG was \emph{Scalar Fallback}. A work-stealing method, each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of starvation because the number of such operations is bounded; the wait will eventually terminate, possibly at the expense of doing redundant work.

@@

We confirmed this queue-like progression experimentally by using atomics to record and print the order in which workgroups exited their lookback phase. This 'ordered exit' behavior differs from that of a \emph{Memcpy} kernel, where all in-flight workgroups typically begin writing immediately after completing reads, often leading to high-concurrency, uncoordinated global writes. In contrast, \emph{DF}'s staggered exit pattern implicitly throttles the number of concurrent writers, reducing memory pressure and improving locality. Workgroups that exit in near-order are more likely to access adjacent memory regions, reducing page thrashing and avoiding premature eviction of active pages. On devices whose memory systems may be insufficient to support highly concurrent writes---such as the 7900XT and Mali-G78---these effects plausibly contribute to scan throughput that exceeds \emph{Memcpy}, despite the added arithmetic. To test this hypothesis, we augmented our \emph{Memcpy} kernel with a simplified lookback routine---excluding the scan itself---and observed a $\sim$10\% throughput increase on the 7900XT. These findings suggest that \emph{Chained Scan}'s computational structure not only enables progress without explicit synchronization, but also improves global memory efficiency by shaping the temporal pattern of writes, effectively translating computational work into emergent memory coordination.

@@

  \item \textbf{Total Fallbacks Initiated:} The total number of fallback operations initiated across all workgroups. Ideally, this number would be zero on devices with FPG\@; because multiple workgroups may perform a fallback on the same blocking tile, this number can exceed the number of workgroups launched. In practice, this count is highly dependent on the fairness of the hardware scheduler. On most devices tested, initiated fallbacks remain modest---typically affecting fewer than 10\% of launched workgroups

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near-100\% fallback rates across all devices—except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} consistently outperforms the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test is limited: it measures the cost of increased fallback operations, but does not fully replicate true blocking behavior. In actual starvation, a blocking workgroup stalls before posting its reduction and cannot proceed until it is granted execution by the scheduler. By contrast, our simulation allows all workgroups to eventually progress and post their inclusive reductions to the spine. Thus, true starvation would be expected to increase lookback lengths—contrary to the slight downward trend observed in our simulation. Finally, starvation leads to effective \emph{occupancy reduction}, which we cannot dynamically simulate: execution resources remain available, but fewer subgroups are scheduled, reducing opportunities for context switching to mitigate latency.
